Modelling the travel time of transit vehicles in \rt through a GTFS-based road network using GPS vehicle locations

### 2019-08-01 anzjs2018/sections/03_models.tex
more changes

% For each particle, the transition function generates a plausible trajectory,
% using system noise parameter $\sigma^2$ which describes 
% how the vehicle's speed changes as a Markov process.
% The vehicle does not travel constantly along the route,
% however, as it needs to service bus stops along the way,
% so the transition function includes 
% stopping probabilities $\boldsymbol\pi = (\pi_1,\ldots,\pi_J)^\top$ at bus stops,
% dwell times $\boldsymbol\tau = (\tau_1,\ldots,\tau_J)^\top$ for passengers to
% board and disembark (conditional on the bus stopping),
% and the minimum dwell time at stops, $\gamma$,
% for the doors to open and close,
% and account for deceleration and acceleration.
% For these parameters, we used constant values for all stops,
% based on those used by \cite{Hans_2015};
% future work will investigate modelling these in real-time similarly to Section~\ref{sec:kf}.

% The transition function algorithm is shown in Algorithm~\ref{fig:algorithm},
% which implements those features discussed above.
% The stop positions are represented by their distance into trip,
% $\boldsymbol{S} = (S_1, \ldots, S_J)^\top$, and
% the $stop\_index$ function returns the index of a stop,
% given a distance travelled and set of stop distances $\boldsymbol{S}$.
% Lines~2--9 are necessary to decrease the degeneration rate of the filter,
% which can occur when there is a long delay between observations
% simply because the bus has not moved due to traffic or other reasons. 
% We therefore place positive probability for some particles to remain stationary.

% Lines~11--12 of Algorithm~\ref{fig:algorithm} implement the system noise,
% which involves changing the vehicle's speed each second (i.e., acceleration).
% Lines~13--25 handle arrival behaviour at stops:
% if the particle passes the next stop,
% it has a chance of stopping (lines~18--19),
% in which case the dwell time is sampled and subtracted from the 
% time remaining to travel (lines~20--21).
% Finally, to track travel time along segments,
% at each point where the vehicle travels (lines 7 and 26),
% the travel time for the current segment, $z_j$, is incremented.
% The mutation ends once the travel time $\delta$ reaches zero,
% or when the vehicle reaches the end of the route (lines 10, 14).
% Once complete, the particle's new state is returned (line 29).

@@

The primary objective of the network model is to estimate the \rt traffic conditions
(travel time) along roads in the transit network, 
as well as make short-term predictions for estimating arrival times.
In this paper, we develop a simple model in which each road is treated independently.
Future work will explore correlations between adjacent roads and using historical data
to improve the model and subsequent predictions.

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$,
and is assumed to be constant over time 
with system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$.
Observations of the state, $Z_c^\ell$, are obtained from the particle filter in Section~\ref{sec:pf},
and have measurement error $e_c^\ell \sim \mathrm{N}(0, R_c^\ell)$ as estimated 
using the uncertainty of particle estimates.
The model from (\ref{eq:rbe_model}) reduces to
\begin{equation*}
\begin{split}
\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^\ell &= \theta_c^\ell + e_c^\ell
\end{split}
\end{equation*}
where $\Delta_c = t_c - t_{c-1}$.

Since multiple vehicles can travel along a road simultaneously,
we used an information filter to implement the network model.
The information filter is a transformation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
This allows multiple observations to be added together to update the state
in a single iteration.

The state is estimated by its mean $\hat \theta_c^\ell = \mathrm{E}(\theta_c^\ell)$
and uncertainty $P_c^\ell = \mathrm{var}(\theta_c^\ell)$,
which are predicted from the previous state estimate using the predictive model
\begin{align*}
\label{eq:kf_transition}
\hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c|c-1} \\
P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + \Delta_c^\ell \nu_c^\ell
\end{align*}

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c = (P_{c|c-1})^{-1}$
and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
The travel time estimate of vehicle $m$ along segment $\ell$
is $\bar z_c^{m\ell}$, with uncertainty $s^{m\ell}_c$,
and is transformed to a measurement information covariance matrix 
$B_c^{m\ell}~=~(s^{m\ell}_c)^{-2}$
and measurement information vector $b_c^{m\ell}~=~\bar z^{m\ell}_c (s^{m\ell}_c)^{-2}$.
The total information is the sum of the information over all $M_c^\ell$ vehicles
that traversed segment $\ell$ since the last update,
so the update equation is
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} b^{m\ell}_{c}.
\end{align*}
The desired parameter estimates are obtained 
by the inverse transformations
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}
\end{equation*}
which can now be used to predict the travel times of upcoming vehicles.

>>>>

@@

The primary objective of the network model is to estimate the \rt traffic conditions
(travel time) along roads in the transit network, 
as well as make short-term predictions for estimating arrival times.
In this paper, we model each road segment independently,
as including correlations simplifies the model to not only a one-dimensional Kalman filter,
but it also means computations can be run in parallel,
greatly increasing the real-time performance of the model.

The network state $\boldsymbol\theta_c = \{\theta_c^\ell\}_{\ell = 1}^L$ is the travel time 
of transit vehicles along road segment $\ell$ at time $t_c$,
of which observations $Z_c^{\ell m}$
are obtained from the particle filter for vehicle $m$ as defined in
(\ref{eq:post_tt}), such that 
(after adding the $m$ superscript to identify unique vehicles), 
\begin{equation*}
Z_c^{\ell m} = \mathrm{E}(p(z_c^{\ell m} | \bY^m_{1:k})).
\end{equation*}
The measurement error is also estimated from the particle filter,
\begin{equation*}
R_c^{\ell m} = \mathrm{Var}(p(z_c^{\ell m} | \bY^m_{1:k})).
\end{equation*}

The model of travel times becomes a simple reduction of (\ref{eq:rbe_model}) 
to the one dimensional case, 
with $f$ and $h$ both unity,
system noise $v_c^\ell \sim \mathrm{N}(0, \nu^2)$
with $\Delta_c = t_c - t_{c-1}$
and $\nu_\ell^2$ is the variance of travel times per second,
and the observation error $e_c^{\ell m} \sim \mathrm{N}(0, R_c^{\ell m})$:
\begin{equation*}
\begin{split}
\theta_c^\ell &= \theta_{c-1}^\ell + \Delta_c^\ell v_c^\ell \\
Z_c^{\ell m} &= \theta_c^\ell + e_c^{\ell m}
\end{split}
\end{equation*}

Since multiple vehicles can travel along a road simultaneously,
we used an information filter to implement the network model.
The information filter is a transformation of the Kalman filter in which the
\emph{information matrix} and \emph{information vector} are used in place of 
the covariance matrix and state vector, respectively.
This allows multiple observations to be added together to update the state
in a single iteration,
which is the case when multiple vehicles travel along a road at the same time.

The state at time $t_{c-1}$ is parameterised by its mean and variance,
conditional on all of the data from all vehicles up until time $t_{c-1}$,
$\hat \theta_{c-1|c-1}^\ell = \mathrm{E}(\theta_c^\ell | Z_{1:c-1}^{\ell\boldsymbol{\cdot}})$
and \mbox{$P_{c-1|c-1}^\ell = \mathrm{var}(\theta_{c-1}^\ell | Z_{1:c-1}^{\ell\boldsymbol{\cdot}})$}, respectively,
which are predicted from the previous state estimate using the predictive model
\begin{align*}
\label{eq:kf_transition}
\hat \theta^\ell_{c|c-1} &= \hat \theta^\ell_{c-1|c-1} \\
P^\ell_{c|c-1} &= P^\ell_{c-1|c-1} + (\Delta_c^\ell \nu_c^\ell)^2
\end{align*}

For the update step, the parameters are transformed into an information
space parameterised by the information matrix $U^\ell_c = (P_{c|c-1})^{-1}$
and the information vector $u^\ell_c~=~\hat \theta^\ell_{c|c-1} (P^\ell_{c|c-1})^{-1}$.
The travel time estimate of vehicle $m$ along segment $\ell$,
$Z_c^{\ell m}$, along with its uncertainty $R^{\ell m}_c$,
are transformed to a measurement information covariance matrix 
$B_c^{\ell m}~=~(R^{\ell m}_c)^{-2}$
and measurement information vector $b_c^{\ell m}~=~Z^{\ell m}_c (R^{\ell m}_c)^{-2}$.
The total information is the sum of the information over all $M_c^\ell$ vehicles
that traversed segment $\ell$ since the last update,
so the update equation is
\begin{align*}
U^\ell_{c|c} &= U^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} B^{m\ell}_{c} \\
\hat u^\ell_{c|c} &= \hat u^\ell_{c|c-1} + \sum_{m=1}^{M_c^\ell} b^{m\ell}_{c}.
\end{align*}
The desired parameter estimates are obtained 
by the inverse transformations
\begin{equation*}
\hat \theta^\ell_{c|c} = \frac{\hat u^\ell_{c|c}}{U^\ell_{c|c}} 
\quad\text{and}\quad
P^\ell_{c|c} = \frac{1}{U^\ell_{c|c}}
\end{equation*}
which can now be used in the prediction of travel times
for upcoming buses.

### 2019-08-01 anzjs2018/sections/04_implementation.tex
more changes

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\epsilon$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
which led us to use the values $\epsilon \in \{1,2,3,5\}$ for GPS error. 
For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to an average vehicle speed variance of between 0.1 and 45~meters per second
over a period of 30~seconds.
For each of combination of these parameter values, including varying $N$,
the following values were calculated:
effective sample size, $N_\text{eff}$;
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the (relative) variance of segment travel time estimates.

>>>>

To evaluate the performance of the model,
the simulation was run using a range of values of GPS error, $\epsilon$,
and system noise, $\sigma^2$.
Figure \ref{fig:dist_to_route} shows the distribution of the distance
between each observation and the route using a nearest point algorithm,
which led us to use the values $\epsilon \in \{1,2,3,5\}$ for GPS error in meters.
For the system noise we used values of $\sigma^2\in \{1e^{-4},1e^{-3},1e^{-2},0.05\}$,
which correspond to an average vehicle speed variance of between 0.1 and 45~meters per second
over a period of 30~seconds.
For each of combination of these parameter values, including varying $N$,
the following values were calculated:
effective sample size, $N_\text{eff}$;
degeneration rate, the percentage of iterations in which the vehicle was lost
(no plausible particles);
and the (relative) variance of segment travel time estimates.
